### 2013-10-28 src2/thesis.tex
update

This thesis focuses on data-parallelism, as opposed to control- or 
task-parallelism. Data parallelism is applicable where the input data can 
be divided into chunks and operated upon independently. Data-parallelism is scalable
meaning that usually as problem size increase, the potential for parallelism increase.

@@

Intel ArBB is part of a movement towards thinking of structure and patterns in 
parallel programming. Other languages and libraries with similar aspirations 
are NVIDIA Thrust~\citet{THRUST} and Data.Array.Accelerate~\citet{ACCELERATEDAMP11}.
In the book ``Structured Parallel Programming'' by Michael McCool et al, a case is 
made for thinking in terms of building blocks for parallel programming~\citet{STRUCTURED}. 
They emphasise the importance of collective array operations such as map, fold, scans, 
permutations and gather/scatter operations. The book contains examples in ArBB, Cilk Plus, 
Intel Threading Building Blocks, OpenMP and OpenCL. Not all these languages include 
the patterns as abstractions, but most are possible to express in each of the languages. 
The patterns, while very powerful as abstractions, also provide tools for thinking about 
parallel algorithms. The ArBB library support flat regular data-parallelism, but also 
irregular and to some degree nested data-parallelism.

@@

\noindent In this example above one block of 32 threads using no shared memory, is launched.

A kernel that uses shared memory takes a form similar to whats shown below.

@@

\noindent A portion of the input array is stored into shared memory. A 
barrier synchronisation is used to ensure that all threads within the block 
has stored its value into shared memory (only necessary if threads will 
communicate across warp boundaries). After computing on the data locally 
it is written back into global memory.

@@

Writing efficient GPU kernels is hard. There are many details that the CUDA 
programmer must be aware of in order to make optimal use of the GPU. 
the NVIDIA ``CUDA C Best Practices Guide''~\citet{BestPrac}, is a good source 
to learn more about these details. Here is a list with explanations of some 
of the practices that the guide deem most important.

@@

\item {\bf Minimise data transfers between host and device:} The bandwidth between device memory 
and GPU is over 20 times of the bandwidth between host memory and the GPU device via the PCIe bus.
This means that, first of all one must evaluate if the computation to be perform is significant 
enough to warrant a transfer to the GPU. If it computation is not arithmetic intensive it may be 
better to do the work on the CPU. However, if the data is already in device memory and 
the operation you are about to perform on it is more efficiently handled by the CPU, it may still 
be better to let the GPU do the work. In other words, compute near the data unless transferring it 
pays off.

\item {\bf Ensure that global memory accesses are coalesced:} Certain access patterns into 
GPU device memory can be {\em coalesced} (combined into few memory transactions). Unfortunately 
what these access patterns are very some depending on the GPU. For a representative model of 
GPU the rule is that concurrent memory accesses (by threads within a warp) will be coalesced 
to as many transactions as cache lines touched (128b l1 cache lines). This means that strided 
accesses from within a warp are not favoured.

\item {\bf Minimise use of global memory:} Accessing global device memory can take as 
many 400 to 600 clock cycles if the data is not cached. The programmer should 
use shared memory as much as possible to avoid redundant loading from global memory.
Shared memory can also be used to coalesce reads from global memory. Shared memory is 
a limited resource in the MP, so if a kernel uses very much shared memory it means 
fewer such blocks can be active on the MP.

\item {\bf Use a multiple of 32 threads per block:} Groups of 32 threads, warps, are the 
scheduled unit on a GPU. If a warp accesses memory (and will be waiting for it arrive during 
600 cycles) it will be swapped out and another warp will take its place. This means that 
the number of available warps decides how well latency can be hidden. A block that is not a 
multiple of 32 leads to wasting GPU resources, some processing elements will stand idle during 
the execution of the not full warps.

\item {\bf Avoid different execution paths within a block:} Any branching instructions that 
causes the execution paths to diverge within a warp will affect performance. When the execution 
paths within a warp diverges the computation is serialised.

\item {\bf Do not use {\tt \_\_synchthreads()} in diverging code:} Extreme care must be taken 
when synchronising within any conditionally executed code; every thread must reach this barrier.
Failing to ensure this will likely lead to the kernel producing incorrect results. Note that 
this is a major limitation to the composability of CUDA functions. If it is safe or 
not to call some function at a certain place in your code, can only be decided by inspecting 
that functions implementation. 
\end{itemize}

The list above provides some very general guide lines of what to think about when 
writing GPU kernels. The lesson is that fast memory near the processing element 
should be preferred over slow memory far away. However, this comes at the effort 
of decomposing computation and data in a way that makes this possible.

@@

This thesis applies embedded language methodology to data-parallel programming. An 
embedded language is implemented as a library in a host language, Haskell in this case. 
The embedded language approach is beneficial since we are not entirely sure how 
to program current and future highly parallel computers. Implementing embedded languages allows 
us more rapid prototyping of ideas. Other benefits of embedded languages often listed 
are include making use of the host language's type system for guaranteeing generating correctly typed 
target code. Already existing libraries in the host language (if they are general enough) 
can be used in the embedded language. The host language can also be used as a powerful macro 
language for your embedded language.

@@

The embedded language approach has become popular to use as a way of raising the 
level of abstraction in fields where the default is to use low-level languages. 
For example, Lava and Wired in the fields of hardware design and 
verification and Feldspar for digital signal processing~\citet{LAVA,Wired,FELDSPAR2010}. For GPU 
programming, there are numerous embedded approaches using various host languages. 
Two languages for GPU programming embedded in Haskell are  Accelerate and Nikola 
\citet{ACCELERATEDAMP11, NIKOLA}. And there is Intel ArBB, embedded in C++, with the 
goal of being parallel across platforms \citet{ARBB2011}.
Related work and information about the tools, libraries and language that make the 
wider context surrounding Obsidian can be found in section~\ref{sec:relatedwork}.

@@

{\tt ForAll} takes a number of parallel iterations, a body represented by a function 
from an index to a program. The body is a {\tt Program} at a some level {\tt t} while the 
resulting program is at a level step above. This means a thread program can be turned into 
a block program by using {\tt ForAll} or that a block program can be turned into a grid program.

@@

The approach taken to embed the scalar language is very similar to what is described 
in ``Compiling Embedded Languages''~\citet{COMPILEEDSL}; but we use a Generalised Algebraic 
Datatype (GADT) to obtain typed expressions. An alternative to the GADT, is to use phantom 
types. Earlier versions of Obsidian (section~\ref{sec:paperB} used phantom types.

@@

Pull arrays are parameterised on both element ({\tt a}) and length ({\tt s}) type. The length 
parameter is used to be able to represent arrays with both static (known at Haskell runtime) 
or dynamic (length represented by an expression, an unknown length). When creating block and 
warp level computations we want to know the exact lengths in order to avoid generating code 
riddled with conditionals concerning array lengths. Conditionals are problematic on GPUs and 
while general programs that run for any size is possible when performance is crucial a fixed 
size is recommended.

One benefit of pull arrays are that they give fusion of all operations on them for free. for example {\tt map f} is implemented on pull arrays by composing {\tt f} with the indexing function:

@@

Now, we see that {\tt map f . map g} is the same as {\tt map f . g} since both result in {\tt f}
and {\tt g} becoming composed onto the indexing function. No intermediate array is built in 
memory.

@@

Push arrays were added to Obsidian as a complement to pull arrays. Pull arrays are 
good because they are so simple to understand and easy to parallelise. However certain
operations on pull arrays yield code that is not suitable for GPU execution. Concatenation 
or interleaving of pull arrays result in conditionals in the indexing function. The 
interleaving case, which is worst, leads to diverging branching in every warp. If branches 
diverge the GPU will execute turn off the processing elements taking one path and 
allow the others to progress; then turning to the other branch. The computation of the 
branches has become serialised and compute resources are wasted.

@@

\begin{verbatim} 
convertToPush :: Pull Word32 e 
               -> Push Block Word32 e 
convertToPush (Pull n ixf) =
    Push n $ \wf -> 
      forAll (fromIntegral n) $ \i -> wf (ixf i) i
\end{verbatim}

The function above takes a pull array and creates a push array with a parallel 
iteration schema over threads within a block. There are other iteration schemas to 
chose from such as sequential, combinbations of sequential and paralell and parallel 
across both threads and blocks (two level parallelism).

Push arrays give many of the same benefits that pull arrays have. Using pull 
arrays we also get fusion of operations as default. The implementation of map 
(shown below) illustrates how maping a function {\tt f} over a push array results 
in an application of {\tt f} in the write-function.

@@

When compiling Obsidian programs to CUDA, we only need to be concerned with the 
{\tt Program} datatype. Given the hierarchy level type parameters to the program datatype 
we know that the AST can be quite directly translated to CUDA. We know that there will 
be at most one level of nestedness in the parallelism (several blocks, each of several threads). 
This means that we do not need to do any transformations on the AST before generating CUDA code. 
Paper F (section~\ref{sec:paperF}) contains technical details of the compilation process.

@@

We have implemented basic algorithms such as sorting, merging, fractal generation, reductions and parallel prefix (scans). Often we have managed to get very good performance, close to that 
of hand optimised CUDA code. However, there are also cases where we just cant seem to reach 
the level of performance that we would want. One such example is parallel prefix, an 
application that has been haunting us from the very beginning. Scan is problematic to us because 
it is best implemented as an in-place algorithm while our array representations do 
not allow for in-place updated (there is no memory associated with them to update). 
We are trying to improve this situation by adding a third array representation to Obsidian,
mutable arrays.

Paper F (section~\ref{sec:paperF}), tells a story about how to implement and tune 
a reduction kernel in Obsidian. At the end the implemented reduction kernel is 
faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
reduction in accelerate corresponds to just one single hand optimised skeleton. We would have 
been happy just getting close.

@@

Paper D (section~\ref{sec:paperD}), introduces push arrays and uses them in the 
implementation of sorting networks to quite good results. With push arrays we 
where able to experiment with more detailed decompositions (what thread computes 
what value) of the algorithms.

@@

EmbArBB provides many of the ArBB operations on Arrays butprovides a Repa like 
shape-polymorphic interface to them. For example a reduction operation can take a two  
dimensional vector and reduce all rows or columns and provide a one dimensional result. Or 
it takes a one dimensional vector and returns a single value (zero dimensional vector).

@@

The {\tt addReduce} function takes an input vector that is atleast one dimensional 
({\tt t:.Int}) and returns a vector that has one less dimension ({\tt t}).

@@

EmbArBB is a traditional deep embedding, there is an AST data type with constructors 
representing each ArBB language feature. It is true that a deep embedding of ArBB 
leads to duplication of effort. The low-level operations exported by the 
ArBB C interface already construct an AST within the ArBB system. One possible benefit 
of the deep embedding is that it buys independence from the ArBB backend. It is possible 
to compile the AST generated on the Haskell side to some other target language.

@@

In paper H ,section~\ref{sec:paperH}, we benchmark EmbArBB against ArBB in C++ and the Haskell 
Repa library. EmbArBB compares favourably to Repa and is often a lot faster. EmbArBB and the ArBB 
C++ performance is not distinguishable from EmbArBB's indicating that our Haskell embedding 
adds no extra overhead. Of course this is expected, since once the ArBB system has generated 
the code, that code is identical no matter if the AST was build via Haskell or C++ calls. 
Missing is a comparison of how well EmbArBB and C++ compares in regards to the time it actually 
takes to go through the process of making those ArBB C API calls. Although if the generated 
code is used many times the caching ArBB does of generated code means that the initial code 
generation cost can be amortised.

@@

In figure~\ref{fig:researchmatrix} our work is placed in relation to other languages,
embedded languages and libraries for parallel programming. The languages and libraries consider 
are data-parallel; languages for control parallelism or distributed message passing system are 
not considered. The systems are roughly divided 
into three groups, based on their level of abstraction. There are low-level languages, here 
I place languages that are imperative and C-like. In the middle layer I place languages 
that go a bit further, they have higher level abstractions and are more easily composable. 
In the highest level I place languages that completely abstract away from details of the 
hardware they run on. The division is not free from personal bias and the placement of languages
in boxes was not always easy.

@@

% ---------------------------------------------------------------------------
% Obsidian, Accelerate, Nikola, Thrust, CUB 
% ---------------------------------------------------------------------------
With Obsidian we try to target the sparsely occupied area of mid-level GPU programming. With 
CUDA~\citet{wwwcuda} and OpenCL~\citet{OpenCL}, the programmer has full control of how to divide 
the computation amongst threads and blocks. Using Accelerate~\citet{ACCELERATEDAMP11}, 
Nikola~\citet{NIKOLA} 
and Thrust~\citet{THRUST} the programmer would use 
high-level patterns without direct insight into how the application would be mapped onto 
threads and blocks of the GPU. Using Obsidian we want to bridge that gap by both being 
more composable than CUDA but give the programmer more control of what the computation will 
look like on the GPU. This is done by allowing the programmer to specify and compose 
thread-level and block-level code. Recently, the CUB library arrived that is built on a 
similar idea, that GPU programmers need to have block/thread level control to get maximum 
performance, while maintaining a higher level and more Thrust-like programmer 
interface~\citet{CUB}. Before CUB, and as far as I am aware, we were alone at targeting 
this level.

% ---------------------------------------------------------------------------
% ArBB EmbArBB
% ---------------------------------------------------------------------------
The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
composes to build his application. I place the ArBB and EmbArBB in a box slightly lower than 
Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
they are not general primitives. Where Accelerate have a single higher order reduction 
operation, ArBB/EmbArBB have {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
%I will go into 
%what benefits there are to having ArBB embedded in Haskell compared to C++ in

% ---------------------------------------------------------------------------
% Microsoft Accelerator
% ---------------------------------------------------------------------------
Microsoft Accelerator~\citet{ACCELERATOR}, is a language embedded in C\#. Accelerator has
a data-parallel arrays and a set of aggregate operations (operations on entire arrays). The 
operations on data-parallel arrays are JIT-compiled to GPU code. Accelerator can also 
compile to multithreaded CPU code and make use of SIMD units of modern 
CPUs~\citet{ACCELERATORCPU}; this makes the Accelerator system related to ArBB.

@@

Repa (Regular Shape-polymorphic Arrays)~\citet{REPA}, is a Haskell library for 
data-parallel programming. 
Repa uses the same array representation as Obsidian, in Repa called a delayed
array, which is a function from index to element. However, in Repa arrays have shape, 
they can be of any dimensionality and there shape-polymorphic functions on these. 
For example {\tt map} can be applied to an array of any shape.

Meta-Repa is a reimplementation of the Repa library using deep embedded language 
techniques. Meta-Repa programs build ASTs that are then compiled (using template 
Haskell) to Haskell code using low-level parts of the Repa library for its parallelism.
The Meta-Repa approach to implementing the Repa library 
gives inlining of operations for free, where Repa programs needs to be annotated 
with pragmas to ensure inlining. On the other hand, in Meta-Repa the programmer 
needs to explicitly state when something should not be inlined (using a {\tt force} primitive).
This is related to our work on EmbArBB that also uses an existing library (ArBB) 
for its parallelism. In reference~\citet{FPCDSL}, a case is made for this style 
of embedding prexisting DSLs.

@@

When we first started working on Obsidian we used Lava~\citet{LavaICFP} as a inspiration. 
We wanted to mimic the Lava programming style but generate GPU code. At first this worked quite 
well and most Obsidian programs superficially looked quite like Lava programs. We have diverged 
from this aspiration since realising that GPUs are quite a different story and performance 
requires more control of GPU architecture specifics.

@@

In paper A~(\ref{sec:paperA}) we describe our method for compiling monadic embedded languages. 
This means we need to take a monadic computation and observe its structure in order to generate
code. This work is closely related to a method described in ``The Constrained Monad Problem''
by Sculthorpe and Gill~\citet{sculthorpe2013constrained}. In this paper they solve a more 
general problem than just compilation but our method is nice because it is easy to understand 
and solves a single problem and does it well. In ``Generic monad constructs'' Persson et al 
also solves the problem of compiling monads using a continuation monad.

@@

Obsidian is a language for GPU kernel implementation. It is not as well integrated 
into Haskell as for example Accelerate. There are ways to execute Obsidian kernels 
from within Haskell; the recommended workflow, however, is to generate the CUDA  
as text and write the host code by hand. This can be seen both a shortcoming 
and a feature. It is hard to integrate Accelerate programs in larger applications 
written in other languages. While when using Obsidian a kernel is generated and 
can be used as part of for example a library.

@@

This thesis contains two approaches to embedding a language for data-parallel programming 
in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
and at little effort from us, very good performance is gained. The performance of 
EmbArBB is entirely due to all the optimisation expertice and effort that invested in 
the ArBB system. We get all that expertice for free. What EmbArBB provide is a 
way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB is based on a 
set of parallel primitives for the programmer to use and compose. Obsidian takes 
a different approach. Built in primitives are very few and very low level, such as 
sequential and parallel for loops. On top of the low-level capabilities a library 
of high-level primitives can be built. The key is that we want the programmer to 
be able to specify how to implement these higher level abstractions.

Looking at the related work, we see that almost all high level GPU programming 
languages provide abstractions of parallel primitives, aggregate operations, 
maps, folds and reductions, giving justification to our investigation of GPU 
programming without these. We believe that the GPU architecture is still volatile 
enough that there be no ``the right'' way of implementing these primitives. As of 
today, the programmer still needs to explore the design space and find a sollution 
that performs well on a particular GPU.

>>>>

This thesis focuses on data-parallelism, as opposed to control- or 
task-parallelism. Data parallelism is applicable where the input data can 
be divided into chunks and operated upon independently. Data-parallelism is scalable,
meaning that the potential for parallelism increases with problem size.

@@

Intel ArBB is part of a movement towards thinking of structure and patterns in 
parallel programming. Other languages and libraries with similar aspirations 
are NVIDIA Thrust~\citet{THRUST} and Data.Array.Accelerate~\citet{ACCELERATEDAMP11}.
In the book ``Structured Parallel Programming'' by Michael McCool et al, a case is 
made for thinking in terms of building blocks for parallel programming~\citet{STRUCTURED}. 
They emphasise the importance of collective array operations such as map, fold, scans, 
permutations and gather/scatter operations. The book contains examples in ArBB, Cilk Plus, 
Intel Threading Building Blocks, OpenMP and OpenCL. Not all these languages include 
the patterns as abstractions, but most are possible to express in each of the languages. 
The patterns, while very powerful as abstractions, also provide tools for thinking about 
parallel algorithms. The ArBB library supports flat regular data-parallelism, but also 
irregular and to some degree nested data-parallelism.

@@

\noindent In the example above one block of 32 threads, using no shared memory, is launched.

A kernel that uses shared memory takes the form:

@@

\noindent A portion of the input array is stored into shared memory. A 
barrier synchronisation is used to ensure that each threads within the block 
has stored its value into shared memory (only necessary if threads will 
communicate across warp boundaries). After computing on the data locally, 
it is written back into global memory.

@@

Writing efficient GPU kernels is hard. There are many details that the CUDA 
programmer must be aware of in order to make optimal use of the GPU. 
The NVIDIA ``CUDA C Best Practices Guide''~\citet{BestPrac}, is a good source 
to learn more about these details. Here is a list, with explanations of some 
of the practices that the guide deems most important.

@@

\item {\bf Minimise data transfers between host and device:} The bandwidth between device memory 
and GPU is much higher than the bandwidth between host memory and the GPU device via the PCIe bus.
This means that, first of all one must evaluate if the computation to be performed is significant 
enough to warrant a transfer to the GPU. If the computation is not arithmetic intensive, it may be 
better to do the work on the CPU. However, if the data is already in device memory and 
the operation that is about to perform on it is more efficiently handled by the CPU, it may still 
be better to let the GPU do the work. In other words, compute near the data unless transferring it 
pays off.

\item {\bf Ensure that global memory accesses are coalesced:} Certain access patterns into 
GPU device memory can be {\em coalesced} (combined into few memory transactions). Unfortunately 
what these access patterns are varies somewhat depending on the GPU. For a typical model of 
GPU, the rule is that concurrent memory accesses (by threads within a warp) will be coalesced 
to as many transactions as cache lines touched (128b l1 cache lines). This means that strided 
accesses from within a warp are not favoured.

\item {\bf Minimise use of global memory:} Accessing global device memory can take as 
many as 400 to 600 clock cycles if the data is not cached. The programmer should 
use shared memory as much as possible to avoid redundant loading from global memory.
Shared memory can also be used to coalesce reads from global memory. Shared memory is 
a limited resource in the MP, so if a kernel uses very much shared memory it means 
fewer such blocks can be active on the MP.

\item {\bf Use a multiple of 32 threads per block:} Groups of 32 threads, warps, are the 
scheduled unit on a GPU. If a warp accesses memory (and will be waiting for it arrive during 
600 cycles) it will be swapped out and another warp will take its place. This means that 
the number of available warps decides how well memory latency can be hidden. A block that is 
not a multiple of 32 leads to wasting GPU resources; some processing elements will stand 
idle during the execution of the not full warps.

\item {\bf Avoid different execution paths within a block:} Any branching instruction that 
causes the execution paths to diverge within a warp will affect performance. When the execution 
paths within a warp diverge, the computation is serialised.

\item {\bf Do not use {\tt \_\_synchthreads()} in diverging code:} Extreme care must be taken 
when synchronising within any conditionally executed code; every thread must reach this barrier.
Failing to ensure this will likely lead to the kernel producing incorrect results. Note that 
this is a major limitation to the composability of CUDA functions. Wheather or not it is safe to 
call some function in a certain place in your code can be decided only by inspecting that function's 
implementation. 
%f it is safe or 
%not to call some function at a certain place in your code, can only be decided by inspecting 
%that functions implementation. 
\end{itemize}

The list above provides some very general guidelines on what to think about when 
writing GPU kernels. The lesson is that fast memory near the processing element 
should be preferred over slow memory far away. However, this comes at the effort 
of decomposing computation and data in a way that makes this possible.

@@

This thesis applies embedded language methodology to data-parallel programming. An 
embedded language is implemented as a library in a host language, Haskell in this case. 
The embedded language approach is beneficial since we are not entirely sure how 
to program current and future highly parallel computers. Implementing embedded languages allows 
us more rapid prototyping of ideas. Other benefits of embedded languages often listed 
are include making use of the host language's type system to guarantee that correctly typed target code is generated. 
Already existing libraries in the host language (if they are general enough) 
can be used in the embedded language. The host language can also be used as a powerful macro 
language for your embedded language.

@@

The embedded language approach has become popular to use as a way of raising the 
level of abstraction in fields where the default is to use low-level languages. 
For example, Lava and Wired~\citet{LAVA,Wired} raises the level of abstraction in the 
fields of hardware design and verification and Feldspar does the same in the area of  digital 
signal processing~\citet{FELDSPAR2010}. For GPU programming, there are numerous 
embedded approaches using various host languages. Two languages for GPU programming 
embedded in Haskell are  Accelerate and Nikola \citet{ACCELERATEDAMP11, NIKOLA}. 
And there is Intel ArBB, embedded in C++, with the goal of being parallel across 
platforms \citet{ARBB2011}. Related work and information about the tools, libraries 
and language that make the wider context surrounding Obsidian can be found in 
section~\ref{sec:relatedwork}.

@@

{\tt ForAll} takes a number of parallel iterations, and a body represented by a function 
from an index to a program. The body is a {\tt Program} at some level {\tt t} while the 
resulting program is a step above. This means that a thread program can be turned into 
a block program by using {\tt ForAll}, or that a block program can be turned into a grid program.

@@

The approach taken to embed the scalar language is very similar to what is described 
in ``Compiling Embedded Languages''~\citet{COMPILEEDSL}; but we use a Generalised Algebraic 
Datatype (GADT) to obtain typed expressions. An alternative to the GADT, is to use phantom 
types. Earlier versions of Obsidian (section~\ref{sec:paperB}) used phantom types.

@@

Pull arrays are parameterised on both element ({\tt a}) and length ({\tt s}) type. The length 
parameter is used to be able to represent arrays with both static (known at Haskell runtime) 
or dynamic (length represented by an expression, an unknown length). When creating block and 
warp level computations, we want to know the exact lengths in order to avoid generating code 
riddled with conditionals concerning array lengths. It is possible to write kernels that 
are not specialised for a certain size. This is accomplished by adding bounds checks. Since 
conditionals  are problematic from a performance point of view, fixed size kernels are recommended 
when performance is crucial. 
%Conditionals are problematic on GPUs and 
%while general programs that run for any size is possible when performance is crucial a fixed 
%size is recommended.

One benefit of pull arrays is that they give fusion of all operations on them for free. 
For example, {\tt map f} is implemented on pull arrays by composing {\tt f} with the indexing function:

@@

Now, we see that \verb!map f . map g! is the same as \verb!map (f . g)! since both result in {\tt f}
and {\tt g} becoming composed onto the indexing function. No intermediate array is built in 
memory.

@@

Push arrays were added to Obsidian as a complement to pull arrays. Pull arrays are 
good because they are so simple to understand and easy to parallelise. However, certain
operations on pull arrays yield code that is not suitable for GPU execution. Concatenation 
or interleaving of pull arrays result in conditionals in the indexing function. The 
interleaving case, which is worst, leads to diverging branching in every warp. If branches 
diverge, the GPU will turn off the processing elements taking one path and 
allow the others to progress; then turning to the other branch. The computation of the 
branches is serialised and compute resources are wasted.

@@

\begin{verbatim} 
convertToPush :: Pull Word32 e 
               -> Push Block Word32 e 
convertToPush (Pull n ixf) =
    Push n $ \wf -> 
      ForAll (fromIntegral n) $ \i -> wf (ixf i) i
\end{verbatim}

The function above takes a pull array and creates a push array with a parallel 
iteration schema over threads within a block. There are other iteration schemas to 
chose from such as sequential, combinbations of sequential and parallel
across both threads and blocks (two level parallelism).

Push arrays give many of the same benefits that pull arrays have. Using push 
arrays we also get fusion of operations as default. The implementation of map 
(shown below) illustrates how maping a function {\tt f} over a push array results 
in an application of {\tt f} in the write-function.

@@

When compiling Obsidian programs to CUDA, we only need to be concerned with the 
{\tt Program} datatype. Given the hierarchy level type parameters to the program datatype, 
we know that the AST can be quite directly translated to CUDA. We know that there will 
be at most one level of nestedness in the parallelism (several blocks, each of several threads). 
This means that we do not need to do any transformations on the AST before generating CUDA code. 
Paper F (section~\ref{sec:paperF}) contains technical details of the compilation process.

@@

We have implemented basic algorithms such as sorting, merging, fractal generation, reductions 
and parallel prefix (scans). Often we have managed to get very good performance, close to that 
of hand optimised CUDA code. However, there are also cases where we just cannot seem to reach 
the level of performance that we would want. One such example is parallel prefix, an 
application that has been haunting us from the very beginning. Scan is problematic to us because 
it is best implemented as an in-place algorithm, while our array representations do 
not allow for in-place update; there is no memory associated with them to update. 
We are trying to improve this situation by adding a third array representation to Obsidian,
mutable arrays.

Paper F (section~\ref{sec:paperF}) tells a story about how to implement and tune 
a reduction kernel in Obsidian. At the end, the implemented reduction kernel is 
faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
reduction in Accelerate corresponds to just one single hand optimised skeleton. We would have 
been happy just getting close.

@@

Paper D (section~\ref{sec:paperD}) introduces push arrays and uses them in the 
implementation of sorting networks to quite good results. With push arrays we 
were able to experiment with more detailed decompositions (what thread computes 
what value) of the algorithms.

@@

EmbArBB provides many of the ArBB operations on vectors but provides a 
shape-polymorphic interface to them. For example, a reduction operation can take a two  
dimensional vector and reduce all rows or columns and provide a one dimensional result. Or 
it takes a one dimensional vector and returns a single value (zero dimensional vector).

@@

The {\tt addReduce} function takes an input vector that is at least one dimensional 
({\tt t:.Int}) and returns a vector that has one less dimension ({\tt t}).

@@

EmbArBB is a traditional deep embedding; there is an AST data type with constructors 
representing each ArBB language feature. It is true that a deep embedding of ArBB 
leads to duplication of effort. The low-level operations exported by the 
ArBB C interface already construct an AST within the ArBB system. One possible benefit 
of the deep embedding is that it buys independence from the ArBB backend. It is possible 
to compile the AST generated on the Haskell side to some other target language.

@@

In paper H, section~\ref{sec:paperH}, we benchmark EmbArBB against ArBB in C++ and the Haskell 
Repa library. EmbArBB compares favourably to Repa and is often a lot faster. EmbArBB and the ArBB 
C++ performance is not distinguishable from EmbArBB's indicating that our Haskell embedding 
adds no extra overhead. Of course this is expected, since once the ArBB system has generated 
the code, that code is identical no matter if the AST was built via Haskell or C++ calls. 
Missing is a comparison of how well EmbArBB and C++ compares in regards to the time it actually 
takes to go through the process of making those ArBB C API calls. However, if the generated 
code is used many times the caching ArBB does of generated code means that the initial code 
generation cost can be amortised.

@@

In figure~\ref{fig:researchmatrix}, our work is placed in relation to other languages,
embedded languages and libraries for parallel programming. The languages and libraries considered 
are data-parallel; languages for control parallelism or distributed message passing systems are 
not considered. The systems are roughly divided 
into three groups, based on their level of abstraction. There are low-level languages; here 
I place languages that are imperative and C-like. In the middle layer I place languages 
that have higher level abstractions and are more easily composable. 
at the highest level, I place languages that completely abstract away from details of the 
hardware they run on. The division is not free from personal bias and the placement of languages
in boxes was not always easy.

@@

% ---------------------------------------------------------------------------
% Obsidian, Accelerate, Nikola, Thrust, CUB 
% ---------------------------------------------------------------------------
With Obsidian, we try to target the sparsely occupied area of mid-level GPU programming. With 
CUDA~\citet{wwwcuda} and OpenCL~\citet{OpenCL}, the programmer has full control of how to divide 
the computation amongst threads and blocks. Using Accelerate~\citet{ACCELERATEDAMP11}, 
Nikola~\citet{NIKOLA} 
and Thrust~\citet{THRUST} the programmer would use 
high-level patterns without direct insight into how the application would be mapped onto 
threads and blocks of the GPU. Using Obsidian, we want to bridge that gap by both being 
more composable than CUDA and giving the programmer control of what the computation will 
look like on the GPU. This is done by allowing the programmer to specify and compose 
thread-level and block-level code. Recently, the CUB library arrived that is built on a 
similar idea, that GPU programmers need to have block/thread level control to get maximum 
performance, while maintaining a higher level and more Thrust-like programmer 
interface~\citet{CUB}. Before CUB, and as far as I am aware, we were alone at targeting 
this level.

% ---------------------------------------------------------------------------
% ArBB EmbArBB
% ---------------------------------------------------------------------------
The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
composes to build his application. I place the ArBB and EmbArBB in a box slightly lower than 
Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
they are not general primitives. Where Accelerate has a single higher order reduction 
operation, ArBB/EmbArBB has {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
%I will go into 
%what benefits there are to having ArBB embedded in Haskell compared to C++ in

% ---------------------------------------------------------------------------
% Microsoft Accelerator
% ---------------------------------------------------------------------------
Microsoft Accelerator~\citet{ACCELERATOR}, is a language embedded in C\#. Accelerator has
data-parallel arrays and a set of aggregate operations (operations on entire arrays). The 
operations on data-parallel arrays are JIT-compiled to GPU code. Accelerator can also 
compile to multithreaded CPU code and make use of SIMD units of modern 
CPUs~\citet{ACCELERATORCPU}; this makes the Accelerator system related to ArBB.

@@

Repa (Regular Shape-polymorphic Arrays)~\citet{REPA}, is a Haskell library for 
data-parallel programming. 
Repa uses the same array representation as Obsidian, in Repa called a delayed
array, which is a function from index to element. However, in Repa arrays have shape; 
they can be of any dimensionality and there are shape-polymorphic functions on these. 
For example, {\tt map} can be applied to an array of any shape.

Meta-Repa~\citet{METAREPA} is a reimplementation of the Repa library using deep embedded language 
techniques. Meta-Repa programs build ASTs that are then compiled (using template 
Haskell) to Haskell code using low-level parts of the Repa library for its parallelism.
The Meta-Repa approach to implementing the Repa library 
gives inlining of operations for free, where Repa programs need to be annotated 
with pragmas to ensure inlining. On the other hand, in Meta-Repa the programmer 
needs to explicitly state when something should not be inlined (using a {\tt force} primitive).
This is related to our work on EmbArBB that also uses an existing library (ArBB) 
for its parallelism. In reference~\citet{FPCDSL}, a case is made for this style 
of embedding prexisting DSLs.

@@

When we first started working on Obsidian, we used Lava~\citet{LavaICFP} as inspiration. 
We wanted to mimic the Lava programming style but generate GPU code. At first this worked quite 
well and most Obsidian programs superficially looked quite like Lava programs. We have diverged 
from this aspiration since realising that GPUs are quite a different story and performance 
requires more control of GPU architecture specifics.

@@

In paper A~(section \ref{sec:paperA}) we describe our method for compiling monadic embedded 
languages. 
This means to take a monadic computation and observe its structure in order to generate
code. This work is closely related to a method described in ``The Constrained Monad Problem''
by Sculthorpe and Gill~\citet{sculthorpe2013constrained}. In that paper they solve a more 
general problem than just compilation, but our method is nice because it is easy to understand 
and solves a single problem and does it well. In ``Generic monad constructs'' Persson et al 
also solves the problem of compiling monads using a continuation monad.

@@

Obsidian is a language for GPU kernel implementation. It is not as well integrated 
into Haskell as for example Accelerate. There are ways to execute Obsidian kernels 
from within Haskell; the recommended workflow, however, is to generate the CUDA  
as text and write the host code by hand. This can be seen as both a shortcoming 
and a feature. It is hard to integrate Accelerate programs in larger applications 
written in other languages; when using Obsidian a kernel is generated and 
can be used as part of for example a library.

@@

This thesis contains two approaches to embedding a language for data-parallel programming 
in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
and at little effort from us, very good performance is gained. The performance of 
EmbArBB is entirely due to all the optimisation expertice and effort that was invested in 
the ArBB system. We get all that expertice for free. What EmbArBB provides is a 
way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB are based on a 
set of parallel primitives for the programmer to use and compose. Obsidian takes 
a different approach. Built in primitives are very few and very low level, such as 
sequential and parallel for loops. On top of the low-level capabilities, a library 
of high-level primitives can be built. The key is that we want the programmer to 
be able to specify how to implement these higher level abstractions.

Looking at the related work, we see that almost all high level GPU programming 
languages provide abstractions of parallel primitives, aggregate operations, 
maps, folds and reductions, giving justification to our investigation of GPU 
programming without these. We believe that the GPU architecture is still volatile 
enough that there be no ``the right'' way of implementing these primitives. As of 
today, the programmer still needs to explore the design space and find a solution 
that performs well on a particular GPU.

Performance of Obsidian kernels is often good and the Obsidian code that they 
are generated from is often shorter than corresponding CUDA programs. Obsidian 
programs differ from CUDA programs by describing whole programs. With this I mean 
that a CUDA programmer needs to take a leap from the single threaded program 
parameterised on thread Id, to what actually will happen when this program is 
run by hundreds or thousands threads at once. The extensive indexing arithmetic 
often present in CUDA programs can make this leap a hard one. Similar programs 
in Obsidian could use operations on whole arrays such as {\tt map} or explicitly 
bounded parallel loops.
